\speculake: Hiding Malware in Speculative Execution

### 2018-12-07 paper/abstract.tex
formatting changes
abstract intro pass

\begin{abstract}
Recently, the Spectre and Meltdown attacks have revealed serious vulnerabilities in modern CPU
designs, allowing an attacker to exfiltrate data from sensitive programs. These
vulnerabilities take advantage of speculative execution to coerce a processor to
perform computation that would otherwise not occur, leaking the resulting
information via side channels to an attacker.

@@

We introduce \emph{\speculake}\footnote{Executable Spectre}, which compiles
arbitrary malicious code into a seemingly-benign payload binary. When a separate
trigger program runs on the same machine, it mistrains the CPU's branch
predictor, causing the payload program to \emph{speculatively} execute code, and
communicating its results back to the rest of the payload program, changes its
behavior.

We study the extent and types of execution that can be performed
speculatively, and demonstrate several computations that can be
performed covertly. In particular, within speculative execution we are able to
decrypt memory using AES-NI instructions at over 5~kbps. Building on this, we
are additionally able to decrypt and interpret
a custom virtual machine language to perform arbitrary computation,
allowing us to create a covert dial-back shell that executes in a few
milliseconds. We also show how our corresponding trigger program
can be a pre-existing benign application already running on the system, and
demonstrate this concept with OpenSSL driven remotely by the attacker as a
trigger program.

%Finally, we describe defenses, but in all honesty, you're probably not going to like
%them.

>>>>

\begin{abstract}
Recently, the Spectre and Meltdown attacks revealed serious vulnerabilities
in modern CPU designs, allowing an attacker to exfiltrate data from sensitive
programs. These vulnerabilities take advantage of speculative execution to
coerce a processor to perform computation that would otherwise not occur,
leaking the resulting information via side channels to an attacker.

@@

We introduce \emph{\speculake}\footnote{Executable Spectre}, which compiles
arbitrary malicious code into a seemingly-benign payload binary. When a separate
trigger program runs on the same machine, it mistrains the CPU's branch
predictor, causing the payload program to \emph{speculatively} execute its malicious
payload, which communicates speculative results back to the rest of the payload program
to change its real-world behavior.

We study the extent and types of execution that can be performed
speculatively, and demonstrate several computations that can be
performed covertly. In particular, within speculative execution we are able to
decrypt memory using AES-NI instructions at over 5~kbps. Building on this, we
decrypt and interpret
a custom virtual machine language to perform arbitrary computation and system
calls in the real world. We demonstrate this with a proof-of-concept dial back shell,
which takes only a few milliseconds to execute after the trigger is issued.
We also show how our corresponding trigger program
can be a pre-existing benign application already running on the system, and
demonstrate this concept with OpenSSL driven remotely by the attacker as a
trigger program.

### 2018-12-07 paper/introduction.tex
Added description of arbitrary memory read methodology
abstract intro pass

Previously, it was assumed that speculative execution results remain invisible
if discarded, as careful CPU design maintains strict separation between
speculative results and updates to architectual state. However, recent research
has revealed side channels that violate this separation, and researchers have
demonstrated ways to exfiltrate results from speculative computation. Most
notably, the Spectre vulnerability allows attackers to leak information from
purposefully mis-speculated branches in a victim process~\cite{spectre}. The
Meltdown vulnerability uses speculative results of an unauthorized memory read
to sidestep page faults and leak protected memory from the
kernel~\cite{meltdown}. Both of these vulnerabilities focus on extracting secret
data from a process or operating system. Recent follow-up work has revealed
other Spectre ``variants'',  including speculative buffer overflows, speculative
store bypass, and using alternative side channels besides the
cache~\cite{kiriansky2018speculative, spectre_ng}. In addition, several attacks
have leveraged Spectre to attack Intel's SGX~\cite{chen2018sgxpectre,
spectre_sgx, foreshadow}, and perform remote leakage
attacks~\cite{schwarz2018netspectre}.

@@

This target location can be in a region that is neither read
nor executed normally by the payload program, making this logic effectively dead code to
any static or dynamic reachability analysis. After the CPU discovers the
mis-speculation at the target location, it will discard the results and continue
executing from the correct destination. However, this still gives the payload program a
limited speculative window where it can perform arbitrary computation, and can
communicate results back to the ``real world'' via a side
channel. Figure~\ref{fig:high-level} shows the variant of \speculake that uses a
trigger program to mis-train the CPU's indirect branch predictor, causing the
payload program to briefly speculatively execute a hidden target function.

@@

It is also possible that a trigger program be a
\emph{benign} program already on the victim's computer. We
show this using the OpenSSL library as a benign trigger
program in Section~\ref{subsec:openssl}, activating a malicious payload program
when an adversary connects to the infected OpenSSL server using a
TLS connection with a specific cipher suite.

@@

Simulating or modelling the speculative execution path is a difficult task for a
program analyst hoping to reverse engineer an \speculake binary. First, the
analyst must reverse engineer and accurately model the closed-source proprietary
components of the target CPU, including the branch predictor, cache hierarchy,
out-of-order execution, and hyperthreading, as well as taking into account the
operating system's process scheduling algorithm. In contrast, the \speculake
author only has to use a partial model of these components and produce binaries
that take advantage of them, while the analyst's model must be complete to
capture all potential \speculake variants. Second, the analyst must run all
potential trigger programs or inputs through the simulator, including benign
programs with real-world inputs. Both of these contribute to a time-consuming and expensive
endeavor for would-be analysts, giving the attacker a significant advantage.

% Results summary
In order to study the potential of \speculake, we implement several example
payload programs and triggers variants, and evaluate their performance. We find
that the payload program's speculative
window is mainly limited by the CPU's reorder buffer, which
allows us to execute between 100 and 200 instructions speculatively on modern
Intel CPUs. While brief, we show how to perform execution in short steps,
communicating intermediate results back to the ``real world'' part of the
payload program. Using this technique, we demonstrate implementing a universal
Turing machine (demonstrating arbitrary computation), a custom instruction set
architecture that fits within the constraints of speculative execution, and show
the ability to perform AES decryption using AES-NI instructions.

>>>>

Previously, it was assumed that speculative execution results remain invisible
if discarded, as careful CPU design maintains strict separation between
speculative results and updates to architectural state. However, recent research
has revealed side channels that violate this separation, and researchers have
demonstrated ways to exfiltrate results from speculative computation. Most
notably, the Spectre vulnerability allows attackers to leak information from
purposefully mis-speculated branches in a victim process~\cite{spectre}. The
Meltdown vulnerability uses speculative results of an unauthorized memory read
to sidestep page faults and leak protected memory from the
kernel~\cite{meltdown}. Both of these vulnerabilities focus on extracting secret
data from a process or operating system. Recent follow-up work has revealed
other Spectre ``variants'',  including speculative buffer overflows, speculative
store bypass, and using alternative side channels besides the
cache~\cite{kiriansky2018speculative, spectre_ng}. In addition, several attacks
have leveraged Spectre to attack Intel's SGX~\cite{chen2018sgxpectre,
spectre_sgx, foreshadow}, and perform remote leakage
attacks~\cite{schwarz2018netspectre}.

@@

This target location can be in a region that is neither read
nor executed normally by the payload program, making this logic effectively dead code to
any static or dynamic reachability analysis. After the CPU discovers the
mis-speculation at the target location, it will discard the results and continue
executing from the correct destination. However, this still gives the payload program a
limited speculative window where it can perform arbitrary computation, and can
communicate results back to the ``real world'' via a side
channel. Figure~\ref{fig:high-level} shows the variant of \speculake that uses a
trigger program to mis-train the CPU's indirect branch predictor, causing the
payload program to briefly execute a hidden target function speculatively.

@@

It is also possible for a trigger program to be a
\emph{benign} program already on the victim's computer. We
show this using the OpenSSL library as a benign trigger
program in Section~\ref{subsec:openssl}, activating a malicious payload program
when an adversary repeatedly connects to the infected OpenSSL server using a
TLS connection with a specific cipher suite.

@@

Simulating or modelling the speculative execution path is a difficult task for a
program analyst hoping to reverse engineer an \speculake binary. First, the
analyst must reverse engineer and accurately model the closed-source proprietary
components of the target CPU, including the branch predictor, cache hierarchy,
out-of-order execution, and hyperthreading, as well as taking into account the
operating system's process scheduling algorithm. In contrast, the \speculake
author only has to use a partial model of these components and produce binaries
that take advantage of them, while the analyst's model must be complete to
capture all potential \speculake variants. Second, the analyst must run all
potential trigger programs or inputs through the simulator, including benign
programs with real world inputs. Both of these contribute to a time-consuming and expensive
endeavor for would-be analysts, giving the attacker a significant advantage.

% Results summary
In order to study the potential of \speculake, we implement several example
payload programs and triggers variants, and evaluate their performance. We find
that the payload program's speculative
window is mainly limited by the CPU's reorder buffer, which
allows us to execute up to 200 instructions speculatively on modern
Intel CPUs. While brief, we show how to perform execution in short steps,
communicating intermediate results back to the ``real world'' part of the
payload program. Using this technique, we demonstrate implementing a universal
Turing machine (demonstrating arbitrary computation), a custom instruction set
architecture that fits within the constraints of speculative execution, and show
the ability to perform AES decryption using AES-NI instructions.
